Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-07-07 text/manuscript-full-KS.qmd
revisions rq1 Chicago

@fig-chicago-maps presents the seven Geomarkets in the Chicago metropolitan area by total population across the three census periods. The maps in the figure show how Chicago Geomarkets reflect the classic urban geographical structure of an inner-city core surrounded by a ring of suburbs as described by @RN4982 [p. 12]. Based on the color gradient scale representing total population density, @fig-chicago-maps indicates northern suburb Geomarkets in the metropolitan area - Chain of Lakes, Northwest Suburbs, Northshore, Evanston and Skokie - experienced relatively little population growth from 1980 through 2020. In contrast, the most significant population increases occured in the Western Suburbs (1.2 Million to 1.6 Million) and the South and Southwest Suburbs (less than 1 Million to 1.3 Million) Geomarkets. The City of Chicago Geomarket, forming the metropolitan area's central core, was the only locality to experience a population decline, from 3.1 Million in 1980 to 2.8 Million in 2020, despite remaining the most populous Geomarket.

These modest population shifts belie substantial change in the racial/ethnic composition within and across Geomarkets over time. @fig-chicago-rq1-race illustrates the racial and ethnic composition of each Geomarket across the three census periods. In 1980, nearly all Geomarkets were predominantly White, with proportions ranging from 83% in the South and Southwest Suburbs to approximately 93% in North Shore. In contrast, the City of Chicago had a substantially more diverse population, with White residents comprising only 46% of its total population. One data caveat is that the following race/ethnicity categories were not available in 1980: Asian, non-Hispanic; Two+ races, non-Hispanic; NHPI, non-Hispanic; AIAN non-Hispanic. By 2020, several Geomarkets that were previously predominantly White shifted to a more heterogeneous racial/ethnic distribution, including Chain of Lakes, Northwest Suburbs, Evanston and Skokie, Western Suburbs, and South and Southwest Suburbs. For many Geomarkets, this balance in racial/ethnic composition was the result of large increases in the Latinx population from 1980 to 2020, which ranged from a 7 percentage point increase for the Evanston and Skokie Geomarket to a 19 percentage point increase for Western Suburbs. Some Geomarkets also experienced a substantial increase in Black populations (South and Southwest Suburbs) and in Asian populations (Northwest Suburbs and Evanston and Skokie). @fig-chicago-rq1-race also indicates that, although the North Shore and City of Chicago Geomarkets underwent some changes in racial and ethnic composition, their overall demographic patterns remained relatively the same across the three census periods. <!-- For instance, North Shore experienced some increases in Hispanic (2% in 1980 to 8% in 2020), Asian Pacific Islander (7% in 2000 to 10% in 2020), and Black (2% in 1980 to 3% in 2020) populations yet remained a predominantly White Geomarket (74% in 2020). City of Chicago also experienced relatively modest declines in its Black population (37% in 1980 to 33% in 2020), alongside increases in its Hispanic (15% in 1980 to 26% in 2020) and Asian Pacific Islander populations (4% in 2000 to 6% in 2020), further substantiating the Geomarket as the most racially/ethnic diverse in the metropolitan area. -->

@@

We explore socioeconomic demographics -- median income, percent living below poverty, and educational attainment -- within and across Geomarkets over time in @fig-chicago-rq1-ses. The primary takeaways are the the City of Chicago -- and to a somewhat lesser degree th South and Southwest Suburbs - stands out as having lower median income, higher poverty rates, and lower BA attainment than other Geomarkets. By contrast, the North Shore stands outas being particularly affluent and highly educated. Overall, median income both within and across Geomarkets in the Chicago metropolitan area remained relatively stable across the three census periods. For instance, in 1980, median household income across the Chicago metropolitan area ranged from a high of \$158,000 in the North Shore Geomarket to a low of \$66,000 in the City of Chicago Geomarket. However, median incomes in all other Geomarkets were closer to the upper thresholds of affluence, such as \$120,000 in the Northwest Suburbs, \$119,000 in Evanston and Skokie, \$108,000 in Chain of Lakes, \$105,000 in the Western Suburbs, and \$99,000 in the South and Southwest Suburbs. While overall income patterns across Geomarkets remained consistent through 2020, a median income increase for the City of Chicago Geomarket (\$79,000) and decrease in the South and Southwest Suburbs Geomarket (\$93,000) narrowed the income gap between the lowest and second-lowest Geomarkets to just \$14,000. @fig-chicago-rq1-ses also shows that from 1980 to 2020, the percentage of residents living in poverty declined in the City of Chicago Geomarket while it increased in all other Geomarkets, resulting in a narrowing of poverty gaps across the Chicago metropolitan area. Similar patterns are evident in educational attainment across Geomarkets. In 1980, the City of Chicago had the lowest percentage of residents with a bachelor's degree or higher (13%). However, by 2020, the Geomarket experienced the largest percentage point increase in college degree attainment, reaching 38%.

@@

```{r fig-chicago-rq1-race, echo = FALSE, fig.align = 'center', fig.cap = "Racial/Ethnic Composition of Chicago Area Geomarkets, 1980-2020", out.width = "100%"}
knitr::include_graphics('./../assets/images/rq1_chicago_race.png')
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)
```

```{r fig-chicago-rq1-ses, echo = FALSE, fig.align = 'center', fig.cap = "Socioeconomic Characteristics of Chicago Area Geomarkets, 1980-2020", out.width = "100%"}
knitr::include_graphics('./../assets/images/rq1_chicago_ses.png')
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)
```

>>>>

@fig-chicago-maps presents the seven Geomarkets in the Chicago metropolitan area by total population across the three census periods. The maps in the figure show how Chicago Geomarkets reflect the classic urban geographical structure of an inner-city core surrounded by a ring of suburbs as described by @RN4982 [p. 12]. Based on the color gradient scale representing total population density, @fig-chicago-maps indicates northern suburb Geomarkets in the metropolitan area - Chain of Lakes, Northwest Suburbs, Northshore, Evanston and Skokie - experienced relatively little population growth from 1980 through 2020. In contrast, the most significant population increases occured in the Western Suburbs (1.0 Million to 1.4 Million) and the South and Southwest Suburbs (less than 1 Million to 1.2 Million) Geomarkets. The City of Chicago Geomarket, forming the metropolitan area's central core, was the only locality to experience a population decline, from 3.5 Million in 1980 to 3.2 Million in 2020, despite remaining the most populous Geomarket.

These modest population shifts belie substantial change in the racial/ethnic composition within and across Geomarkets over time. @fig-chicago-rq1-race illustrates the racial and ethnic composition of each Geomarket across the three census periods. In 1980, nearly all Geomarkets were predominantly White, with proportions ranging from 84% in the South and Southwest Suburbs to approximately 96% in North Shore. In contrast, the City of Chicago had a substantially more diverse population, with White residents comprising only 52% of its total population. One data caveat is that the following race/ethnicity categories were not available in 1980: Asian, non-Hispanic; Two+ races, non-Hispanic; NHPI, non-Hispanic; AIAN non-Hispanic. By 2020, several Geomarkets that were previously predominantly White shifted to a more heterogeneous racial/ethnic distribution, including Chain of Lakes, Northwest Suburbs, Evanston and Skokie, Western Suburbs, and South and Southwest Suburbs. For many Geomarkets, this balance in racial/ethnic composition was the result of large increases in the Latinx population from 1980 to 2020, which ranged from a 7 percentage point increase for the Evanston and Skokie Geomarket to a 20 percentage point increase for Chain of Lakes. Some Geomarkets also experienced a substantial increase in Black populations (South and Southwest Suburbs) and in Asian populations (Northwest Suburbs and Evanston and Skokie). @fig-chicago-rq1-race also indicates that, although the North Shore Geomarket underwent some changes in racial and ethnic composition, its overall demographic patterns remained relatively the same across the three census periods. City of Chicago experienced relatively modest declines in its Black population (35% in 1980 to 26% in 2020), alongside increases in its Hispanic (13% in 1980 to 30% in 2020) and Asian Pacific Islander populations (4% in 2000 to 7% in 2020), further substantiating the Geomarket as the most racially/ethnic diverse in the metropolitan area. <!-- For instance, North Shore experienced some increases in Hispanic (2% in 1980 to 8% in 2020), Asian Pacific Islander (7% in 2000 to 10% in 2020), and Black (2% in 1980 to 3% in 2020) populations yet remained a predominantly White Geomarket (74% in 2020). City of Chicago also experienced relatively modest declines in its Black population (37% in 1980 to 33% in 2020), alongside increases in its Hispanic (15% in 1980 to 26% in 2020) and Asian Pacific Islander populations (4% in 2000 to 6% in 2020), further substantiating the Geomarket as the most racially/ethnic diverse in the metropolitan area. -->

@@

We explore socioeconomic demographics -- median income, percent living below poverty, and educational attainment -- within and across Geomarkets over time in @fig-chicago-rq1-ses. The primary takeaways are the City of Chicago -- and to a somewhat lesser degree th South and Southwest Suburbs - stands out as having lower median income, higher poverty rates, and lower BA attainment than other Geomarkets. By contrast, the North Shore stands out as being particularly affluent and highly educated. Overall, median income both within and across Geomarkets in the Chicago metropolitan area remained relatively stable across the three census periods. For instance, in 1980, median household income across the Chicago metropolitan area ranged from a high of \$168,000 in the North Shore Geomarket to a low of \$68,000 in the City of Chicago Geomarket. However, median incomes in all other Geomarkets were closer to the upper thresholds of affluence, such as \$131,000 in Evanston and Skokie, \$120,000 in the Northwest Suburbs, \$111,000 in the Western Suburbs, \$109,000 in Chain of Lakes, and \$99,000 in the South and Southwest Suburbs. While overall income patterns across Geomarkets remained consistent through 2020, a median income increase for the City of Chicago Geomarket (\$80,000) and decrease in the South and Southwest Suburbs Geomarket (\$91,000) narrowed the income gap between the lowest and second-lowest Geomarkets to just \$11,000. @fig-chicago-rq1-ses also shows that from 1980 to 2020, the percentage of residents living in poverty declined in the City of Chicago Geomarket while it increased in all other Geomarkets, resulting in a narrowing of poverty gaps across the Chicago metropolitan area. Similar patterns are evident in educational attainment across Geomarkets. In 1980, the City of Chicago and South and Southwest Suburbabs had the lowest percentage of residents with a bachelor's degree or higher (14%). However, by 2020, the City of Chicago experienced the largest percentage point increase in college degree attainment, reaching 34%.

@@

```{r fig-chicago-rq1-race, echo = FALSE, fig.align = 'center', fig.cap = "Racial/Ethnic Composition of Chicago Area Geomarkets, 1980-2020", out.width = "100%"}
knitr::include_graphics('./../results/graphs/rq1/rq1_chicago_race.png')
#knitr::include_graphics('./../assets/images/rq1_chicago_race.png') [OLD RESULTS BASED ON ZIP GEOMARKETS]
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)
```

```{r fig-chicago-rq1-ses, echo = FALSE, fig.align = 'center', fig.cap = "Socioeconomic Characteristics of Chicago Area Geomarkets, 1980-2020", out.width = "100%"}
knitr::include_graphics('./../results/graphs/rq1/rq1_chicago_ses.png')
#knitr::include_graphics('./../assets/images/rq1_chicago_ses.png') [OLD RESULTS BASED ON ZIP GEOMARKETS]
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)
```
